---
title: "Microbial communities of bat bugs and swallow bugs (DNA and RNA metagenomics)"
output: html_notebook
---

Downstream analysis of the Kraken2/Bracken count tables built in step 4
(`data/bracken_phylum_counts.csv`, `data/bracken_species_counts.csv`).
Twelve profiles: 2 hosts (bat bug *Stricticimex parvus*, swallow bug *Paracimex avium*)
x 3 life stages (female, male, nymph; one pooled sample each) x 2 sequencing types (DNA, RNA).

Because each host x stage x sequence type is a single pooled profile, the statistics
below describe structure in the dataset; they are not replicated tests of biological groups.

```{r setup}
## Run from the repository root: all paths below are relative to it
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

```{r libraries}
library(phyloseq)
library(vegan)
library(ggplot2)
library(tidyr)
library(dplyr)
library(RColorBrewer)

packageVersion("phyloseq")
packageVersion("vegan")
```

```{r settings}
outdir = 'results/'
dir.create(outdir, recursive = TRUE, showWarnings = FALSE)
set.seed(2024)   # rarefaction subsamples at random: the seed makes reruns identical

host_labels = c(bat_bug = 'Bat bug', swallow_bug = 'Swallow bug')
```

```{r import-data}
metaTab = read.csv('data/metadata.csv', check.names = F, stringsAsFactors = F)
rownames(metaTab) = metaTab$sample_id
metaTab$bugs_stage = factor(metaTab$bugs_stage, levels = c('female', 'male', 'nymph'))
metaTab$group = paste(host_labels[metaTab$bugs_host], metaTab$bugs_stage)

read.counts <- function(file) {
    tab = read.csv(file, check.names = F, stringsAsFactors = F)
    mat = as.matrix(tab[, -1])
    rownames(mat) = tab$taxon
    storage.mode(mat) = 'numeric'
    mat = round(mat)
    ## keep the samples described in the metadata, in the metadata order
    missing = setdiff(metaTab$sample_id, colnames(mat))
    if (length(missing) > 0) stop('Samples missing from ', file, ': ', paste(missing, collapse = ', '))
    mat[, metaTab$sample_id]
}

taxMatP = read.counts('data/bracken_phylum_counts.csv')
taxMatS = read.counts('data/bracken_species_counts.csv')

## classified reads per sample (Bracken-estimated)
totalCount = data.frame(sample_id = colnames(taxMatS),
                        phylum_reads = colSums(taxMatP),
                        species_reads = colSums(taxMatS))
write.csv(totalCount, paste0(outdir, 'Table-classified-read-counts.csv'), row.names = F)
totalCount
```

```{r create-phyloseq-objects}
make.phyloseq <- function(mat, rank) {
    ann = matrix(rownames(mat), ncol = 1, dimnames = list(rownames(mat), rank))
    phyloseq(otu_table(mat, taxa_are_rows = T), tax_table(ann), sample_data(metaTab))
}
plsP = make.phyloseq(taxMatP, 'Phylum')
plsS = make.phyloseq(taxMatS, 'Species')
plsS
```

## Sequencing depth

```{r rarefaction-curves}
pdf(paste0(outdir, 'Fig-S-rarefaction-curves-species.pdf'), width = 10, height = 5)
par(mfrow = c(1, 2))
for (seqType in c('DNA', 'RNA')) {
    ids = metaTab$sample_id[metaTab$sequence == seqType]
    rarecurve(t(taxMatS[, ids]), step = max(1, floor(min(colSums(taxMatS[, ids])) / 50)),
              label = T, cex = 0.6, xlab = 'Classified reads', ylab = 'Species',
              main = paste(seqType, '-derived profiles'))
}
dev.off()
```

## Rarefaction and relative abundance

```{r rarefy}
## all 12 profiles are rarefied to the smallest library, so DNA and RNA can be compared
minDepth = min(colSums(taxMatS))
minDepth
rTaxMatS = t(rrarefy(t(taxMatS), minDepth))

calc.relative.abundance <- function(taxMat) {
    taxRel = taxMat
    for (i in 1:ncol(taxMat)) {
        taxRel[, i] = 100 * taxMat[, i] / sum(taxMat[, i])
    }
    taxRel
}
relMatP = calc.relative.abundance(taxMatP)
relMatS = calc.relative.abundance(taxMatS)
```

## Alpha diversity

```{r alpha-diversity}
shannonValue  = diversity(rTaxMatS, index = 'shannon', MARGIN = 2)
observedValue = specnumber(t(rTaxMatS))
pielouValue   = shannonValue / log(observedValue)

alpDat = data.frame(sample_id = metaTab$sample_id,
                    shannon = shannonValue[metaTab$sample_id],
                    observed_species = observedValue[metaTab$sample_id],
                    pielou = pielouValue[metaTab$sample_id],
                    bugs_host = host_labels[metaTab$bugs_host],
                    bugs_stage = metaTab$bugs_stage,
                    sequence = metaTab$sequence)
write.csv(alpDat, paste0(outdir, 'Table-alpha-diversity.csv'), row.names = F)

alpDatl = gather(alpDat, index, value, shannon:pielou)
alpDatl$index = factor(alpDatl$index, levels = c('observed_species', 'shannon', 'pielou'),
                       labels = c('Observed species', 'Shannon', 'Pielou evenness'))

p = ggplot(alpDatl, aes(x = bugs_stage, y = value, fill = sequence)) +
    geom_col(position = position_dodge(width = 0.8), width = 0.7) +
    facet_grid(index ~ bugs_host, scales = 'free_y') +
    scale_fill_manual(values = c(DNA = '#1B9E77', RNA = '#D95F02')) +
    labs(x = 'Life stage', y = '', fill = 'Sequencing') +
    theme_bw()
p
ggsave(paste0(outdir, 'Fig-alpha-diversity.png'), p, width = 7, height = 6, dpi = 300)
```

## Beta diversity

```{r pcoa-per-sequence-type}
## PCoA on Bray-Curtis dissimilarity, separately for DNA- and RNA-derived profiles
for (seqType in c('DNA', 'RNA')) {
    ids  = metaTab$sample_id[metaTab$sequence == seqType]
    bray = vegdist(t(rTaxMatS[, ids]), method = 'bray')
    pcoa = cmdscale(bray, k = 2, eig = T)
    varExp = round(100 * pcoa$eig[1:2] / sum(pcoa$eig[pcoa$eig > 0]), 1)
    ordDat = data.frame(sample_id = ids, PC1 = pcoa$points[, 1], PC2 = pcoa$points[, 2],
                        bugs_host = host_labels[metaTab[ids, 'bugs_host']],
                        bugs_stage = metaTab[ids, 'bugs_stage'])
    p = ggplot(ordDat, aes(PC1, PC2, colour = bugs_host, shape = bugs_stage)) +
        geom_point(size = 4) +
        labs(title = paste(seqType, '-derived profiles (Bray-Curtis PCoA)'),
             x = paste0('PCo1 (', varExp[1], '%)'), y = paste0('PCo2 (', varExp[2], '%)'),
             colour = 'Host', shape = 'Life stage') +
        theme_bw()
    print(p)
    ggsave(paste0(outdir, 'Fig-PCoA-bray-curtis-', seqType, '.png'), p, width = 6, height = 4.5, dpi = 300)
}
```

```{r nmds-combined}
## NMDS on the combined DNA and RNA profiles
brayAll = vegdist(t(rTaxMatS), method = 'bray')
nmds = metaMDS(brayAll, k = 2, trymax = 100, trace = F)
nmds$stress

nmdsDat = data.frame(sample_id = metaTab$sample_id,
                     NMDS1 = nmds$points[, 1], NMDS2 = nmds$points[, 2],
                     bugs_host = host_labels[metaTab$bugs_host],
                     bugs_stage = metaTab$bugs_stage,
                     sequence = metaTab$sequence)
p = ggplot(nmdsDat, aes(NMDS1, NMDS2, colour = bugs_host, shape = sequence)) +
    geom_point(size = 4) +
    geom_text(aes(label = bugs_stage), vjust = -1.2, size = 3, show.legend = F) +
    labs(title = paste0('Bray-Curtis NMDS (stress = ', round(nmds$stress, 3), ')'),
         colour = 'Host', shape = 'Sequencing') +
    theme_bw()
p
ggsave(paste0(outdir, 'Fig-NMDS-bray-curtis-DNA-RNA.png'), p, width = 6.5, height = 5, dpi = 300)
```

```{r permanova}
## PERMANOVA: effects of host species, life stage and sequencing type
perm = adonis2(brayAll ~ bugs_host + bugs_stage + sequence,
               data = metaTab, permutations = 9999, by = 'margin')
perm
write.csv(as.data.frame(perm), paste0(outdir, 'Table-PERMANOVA.csv'))
```

## Taxonomic composition

```{r relative-abundance-plots}
## taxa reaching more than 1% relative abundance in at least one profile; the rest = Others
plot.composition <- function(relMat, rankName, file) {
    keep = rownames(relMat)[apply(relMat, 1, max) > 1]
    selRel = rbind(relMat[keep, , drop = F], Others = 100 - colSums(relMat[keep, , drop = F]))
    dat = data.frame(taxon = rownames(selRel), selRel, check.names = F)
    datl = gather(dat, sample_id, abundance, -taxon)
    datl$bugs_host  = host_labels[metaTab[datl$sample_id, 'bugs_host']]
    datl$bugs_stage = metaTab[datl$sample_id, 'bugs_stage']
    datl$sequence   = metaTab[datl$sample_id, 'sequence']
    datl$taxon = factor(datl$taxon, levels = c(keep[order(rowSums(relMat[keep, , drop = F]), decreasing = T)], 'Others'))

    n = nlevels(datl$taxon)
    palette = c(colorRampPalette(brewer.pal(12, 'Paired'))(n - 1), 'grey80')
    p = ggplot(datl, aes(x = bugs_stage, y = abundance, fill = taxon)) +
        geom_bar(position = 'stack', stat = 'identity') +
        scale_fill_manual(values = palette) +
        facet_grid(~ bugs_host * sequence) +
        labs(x = 'Life stage', y = 'Relative abundance (%)', fill = rankName) +
        theme_bw() +
        theme(legend.key.size = unit(4, 'mm'), legend.text = element_text(size = 7)) +
        guides(fill = guide_legend(ncol = 1))
    print(p)
    ggsave(paste0(outdir, file), p, width = 11, height = 6, dpi = 300)
    write.csv(selRel, paste0(outdir, sub('\\.png$', '.csv', file)))
}

plot.composition(relMatP, 'Phylum',  'Fig-relative-abundance-phylum.png')
plot.composition(relMatS, 'Species', 'Fig-relative-abundance-species.png')
```

## LEfSe (mothur)

LEfSe was run in mothur (v1.44.3) with default parameters (Kruskal-Wallis alpha 0.05,
LDA cut-off 2.0) on the species-level profiles. This chunk writes the mothur input files;
the mothur commands are printed at the end.

```{r export-for-lefse}
lefseDir = paste0(outdir, 'lefse/')
dir.create(lefseDir, showWarnings = F)

## mothur needs simple feature names: species are renamed Otu0001, Otu0002, ... and the key is saved
otuNames = sprintf('Otu%04d', seq_len(nrow(taxMatS)))
write.csv(data.frame(otu = otuNames, species = rownames(taxMatS)),
          paste0(lefseDir, 'otu_to_species.csv'), row.names = F)

write.shared <- function(mat, file) {
    shared = data.frame(label = 'species', Group = colnames(mat), numOtus = nrow(mat),
                        t(mat), check.names = F)
    colnames(shared)[-(1:3)] = otuNames
    write.table(shared, file, sep = '\t', quote = F, row.names = F)
}
write.shared(taxMatS, paste0(lefseDir, 'all.shared'))

## design files: sample -> group, one per comparison
write.design <- function(ids, groups, file) {
    write.table(data.frame(group = ids, treatment = groups), file,
                sep = '\t', quote = F, row.names = F, col.names = F)
}
write.design(metaTab$sample_id, metaTab$bugs_host,  paste0(lefseDir, 'host.design'))
write.design(metaTab$sample_id, metaTab$bugs_stage, paste0(lefseDir, 'stage.design'))
for (h in names(host_labels)) {
    ids = metaTab$sample_id[metaTab$bugs_host == h]
    write.shared(taxMatS[, ids], paste0(lefseDir, h, '.shared'))
    write.design(ids, metaTab[ids, 'sequence'], paste0(lefseDir, h, '_DNA_vs_RNA.design'))
}

cat('Run in mothur (from', lefseDir, '):\n',
    'lefse(shared=all.shared, design=host.design)\n',
    'lefse(shared=all.shared, design=stage.design)\n',
    'lefse(shared=bat_bug.shared, design=bat_bug_DNA_vs_RNA.design)\n',
    'lefse(shared=swallow_bug.shared, design=swallow_bug_DNA_vs_RNA.design)\n')
```

```{r plot-lefse-results}
## after running mothur: read each *.lefse_summary and plot significant species by LDA score
key = read.csv(paste0(lefseDir, 'otu_to_species.csv'), stringsAsFactors = F)
summaries = list.files(lefseDir, pattern = 'lefse_summary$', full.names = T)

for (f in summaries) {
    lef = read.delim(f, stringsAsFactors = F)
    lef = lef[!is.na(lef$LDA) & lef$LDA >= 2 & !is.na(lef$pValue) & lef$pValue < 0.05, ]
    if (nrow(lef) == 0) next
    lef$species = key$species[match(lef$OTU, key$otu)]
    lef = lef[order(lef$Class, lef$LDA), ]
    lef$species = factor(lef$species, levels = unique(lef$species))
    p = ggplot(lef, aes(x = species, y = LDA, fill = Class)) +
        geom_col() + coord_flip() +
        labs(x = '', y = 'LDA score (log10)', title = basename(f)) +
        theme_bw() + theme(axis.text.y = element_text(size = 7, face = 'italic'))
    print(p)
    ggsave(paste0(lefseDir, sub('lefse_summary$', 'png', basename(f))), p,
           width = 7, height = max(3, 0.18 * nrow(lef) + 1.5), dpi = 300)
}
```
